First steps in synthetic algebraic geometry

### 2017-03-14 snippets/open-subfunctors.tex
Comment on wrong characterization

\begin{prop}\label{prop:characterization-open-subfunctor}
Let~$X \in \Zar(S)$ be a Zariski sheaf. A subfunctor~$U
\hookrightarrow X$ is open if and only if any of the following conditions,
formulated in the internal language of~$\Zar(S)$, is
satisfied.
\begin{enumerate}
\item For all~$x \? X$ there is an ideal~$J \subseteq \affl$ such
that~$\affl/J$ is quasicoherent and such that~$x \in U$ if and only if~$1 \in
J$. In formal language:
\begin{multline*}
  \qquad\qquad\Zar(S) \models
  \forall x\?X\_
  \exists J \subseteq \affl\_ \\
  \speak{$J$ is an ideal} \wedge
  \speak{$\affl/J$ is quasicoherent} \wedge
  (x \in U \Leftrightarrow 1 \in J). \end{multline*}
\item For all~$x \? X$ the quotient~$\affl/K$ modulo the radical ideal
\[ K \defeq \{ f\?\affl \,|\, \speak{$f$ \inv} \Rightarrow (x \in U) \}
\subseteq \affl \]
is quasicoherent.
\end{enumerate}
\end{prop}

The displayed conditions are only meaningful in an intuitionistic context as
provided by the big Zariski topos. For instance, the first condition would be
trivially satisfied in classical logic, for any subfunctor~$U \hookrightarrow
X$: Classically, we have~$(x \in U) \vee (x \not\in U)$. If~$x \in U$, we can
pick~$J = (1)$, and if~$x \not\in U$, we can pick~$J = (0)$ (such that the
quotient~$\affl/J$ is isomorphic to~$\affl$, thus finitely presented and
therefore in particular quasicoherent).\footnote{Strictly speaking,
incompability with classical logical surfaces even earlier: in our internal
quasicoherence condition. The map~$E \otimes_\affl A \to [\Spec(A), E]$ which
the condition demands to be bijective has hardly any chance to be surjective if
the law of excluded middle is available to define maps~$\Spec(A) \to E$ by case
distinction.}

@@

\begin{proof}[Proof of Proposition~\ref{prop:characterization-open-subfunctor}]

\end{proof}

>>>>

A basic concept in the functor-of-points approach to algebraic geometry is the
concept of an \emph{open subfunctor}. It is used to delimit schemes from more
general kinds of spaces: A functor is deemed to be a scheme if and only if it
admits a covering by open subfunctors which are representable.

The following definition is phrased in such a way as to apply to any of the
several ways to define the big Zariski topos~$\Zar(S)$. In particular, it
applies to the definition using the site consisting of affine schemes which are
locally of finite presentation over~$S$. If~$S$ is affine, the definition only
refers to affine schemes and open subschemes of affine schemes and is therefore
suitable if one wants to found the theory of schemes using the functorial
approach.

\begin{defn}[{\cite[Définition~I.1.3.6 on page~10]{demazure:gabriel},
\cite[Tag~01JI]{stacks-project}}]
A subfunctor~$U \hookrightarrow X$ in~$\Zar(S)$ is an \emph{open subfunctor} if
and only if for any object~$(T \to S)$ of the site used to define~$\Zar(S)$
and any~$x \in X(T)$ there exists an open subscheme~$T_0 \subseteq T$
such that for any object~$(T' \xra{f}{\to} T \to S)$ of the site used to
define~$\Zar(S)$ the map~$T' \to T$ factors over~$T_0$ if and only if~$X(f)(x)
\in U(T')$.
\end{defn}

\begin{prop}\label{prop:characterization-open-subfunctor}
Let~$X \in \Zar(S)$ be a Zariski sheaf. A subfunctor~$U
\hookrightarrow X$ is open if and only if, from the internal point of view
of~$\Zar(S)$, it holds that for all~$x \? X$ there is an ideal~$J \subseteq
\affl$ such that~$\affl/J$ is quasicoherent and such that~$x \in U$ if and only
if~$1 \in J$. In formal language:
\begin{multline*}
  \qquad\qquad\Zar(S) \models
  \forall x\?X\_
  \exists J \subseteq \affl\_ \\
  \speak{$J$ is an ideal} \wedge
  \speak{$\affl/J$ is quasicoherent} \wedge
  (x \in U \Leftrightarrow 1 \in J). \end{multline*}
\end{prop}

The displayed condition is only meaningful in an intuitionistic context as
provided by the big Zariski topos. In classical logic, the condition would be
trivially satisfied for any subfunctor~$U \hookrightarrow X$: Classically, we
have~$(x \in U) \vee (x \not\in U)$. If~$x \in U$, we can pick~$J = (1)$, and
if~$x \not\in U$, we can pick~$J = (0)$ (such that the quotient~$\affl/J$ is
isomorphic to~$\affl$, thus finitely presented and therefore in particular
quasicoherent).\footnote{Strictly speaking, incompability with classical
logical surfaces even earlier: in our internal quasicoherence condition. The
map~$E \otimes_\affl A \to [\Spec(A), E]$ which the condition demands to be
bijective has hardly any chance to be surjective if the law of excluded middle
is available to define maps~$\Spec(A) \to E$ by case distinction.}

@@

\begin{proof}[Proof of Proposition~\ref{prop:characterization-open-subfunctor}]
...
\end{proof}

\begin{rem}The radical~$\sqrt{J}$ of the ideal~$J$ appearing in
Proposition~\ref{prop:characterization-open-subfunctor} is unique:
It is equal to the radical ideal
\[ K \defeq \{ f\?\affl \,|\, \speak{$f$ \inv} \Rightarrow (x \in U) \}
\subseteq \affl. \]
It's obvious that~$J \subseteq K$ and therefore~$\sqrt{J} \subseteq K$.
For the converse direction, let~$f \in K$ be given. Since~$\affl/J$ is
quasicoherent, the canonical map
\[ (\affl/J)[f^{-1}] \longrightarrow [ \Spec(\affl[f^{-1}]), \affl/J ] \]
is bijective. Since~$\Spec(\affl[f^{-1}]) \cong \brak{\speak{$f$ \inv}}$, the image
of~$1$ is zero: If~$\Spec(\affl[f^{-1}])$ is inhabited, $f$ is invertible and
therefore~$x$ is an element of~$U$. This implies that~$1 \in J$.
Thereby~$\affl/J = 0$. By injectivity of the canonical map, the
algebra~$(\affl/J)[f^{-1}]$ is zero. Therefore~$f^n \in J$ for some natural
number~$n$.\end{rem}

\begin{rem}In view of the previous remark, one cannot hope to simplify the
condition in Proposition~\ref{prop:characterization-open-subfunctor} to read as
follows: ``For any~$x\?X$, the quotient~$\affl/K$ modulo the ideal~$K = \{
f\?\affl \,|\, \speak{$f$ \inv} \Rightarrow (x \in U) \}$ is quasicoherent.
This statement implies the condition in the proposition, but the converse
direction does not hold, since~$\affl/K \cong \affl/\sqrt{J}$ might fail to be
quasicoherent. For instance that's the case if~$U = \emptyset$; then~$K =
\sqrt{(0)}$ is not quasicoherent by Remark~\ref{rem:radical-not-qcoh}.\end{rem}
